---
title: "NYPD Incidents: race by race"
author: "anonymous"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

## Libraries used 
library(readr) 
library(dplyr)
library(lubridate)
library(ggmap)
library(ggplot2)
library(RColorBrewer)
library(stats)


## read in dataset
#loading in pop for global
url_pop = 'https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD'
base_nypd_data = read_csv(url_pop)

#understand how much data is missing 
missing_counts <- base_nypd_data %>%  summarise(across(everything(), ~sum(is.na(.))))

knitr::kable(head(missing_counts))

#we have 59 records with missing geo data, remove
nypd_data <- base_nypd_data %>%
  filter(!is.na(Longitude))

# Convert 'Latitude' and 'Longitude' to numeric if they are factors or characters
nypd_data$Latitude <- as.numeric(as.character(nypd_data$Latitude))
nypd_data$Longitude <- as.numeric(as.character(nypd_data$Longitude))

# Filter out rows where perpetrator race is NA or 'UNKNOWN'
nypd_data_filtered <- nypd_data %>%
  filter(!is.na(PERP_RACE) & PERP_RACE != 'UNKNOWN')

# Registering the API key for Google Maps (replace 'your_api_key' with your actual API key)
register_google(key = "AIzaSyC6Jw4MwQoFPEcZNruGhgrIOPeboczR4uo")

# Get the map
new_york_map <- get_map(location = 'New York', zoom = 10)




# Creating a heatmap of incidents by perpetrator race vs. victim race
race_data <- nypd_data %>%
  mutate(PERP_RACE = ifelse(PERP_RACE %in% c("WHITE HISPANIC", "BLACK HISPANIC"), "HISPANIC - MERGED", PERP_RACE),
         VIC_RACE = ifelse(VIC_RACE %in% c("WHITE HISPANIC", "BLACK HISPANIC"), "HISPANIC - MERGED", VIC_RACE)) %>%
  filter(PERP_RACE != "(null)", PERP_RACE != "", PERP_RACE != "NA", !is.na(PERP_RACE),
         VIC_RACE != "(null)", VIC_RACE != "", VIC_RACE != "NA", !is.na(VIC_RACE)) %>%
  count(PERP_RACE, VIC_RACE)
```


## Data Transformations

- **Removed missing longitude/latitude data** (59 rows).
- **Removed missing race data** (9,300 records). _Note: Acknowledge this may be biasing the overall location data as missing race can have correlation with a number of factors._
- **Consolidated 'Hispanic White' and 'Hispanic Black' into 'Hispanic'.**

## Analysis

## First Figure

- **The first figure takes a broad look into perpetrator vs. victim race to see how incidents map across races.**
- **Incident heatmap is centered around black and hispanic perpetrators with black / hispanic victims**
- **While black on black violence is the highest count, other categories of same race-on-race do not suggest that race-on-race crime is dominant.** 

```{r race_by_race_heatmap, echo=FALSE, fig.cap="Heatmap: Race-on-Race"}
ggplot(race_data, aes(x = PERP_RACE, y = VIC_RACE, fill = n)) +
  geom_tile(color = "white") +  # Add white lines to distinguish the tiles
  geom_text(aes(label = n), color = "black", size = 3) +  # Add incident count labels
  scale_fill_gradient(low = "blue", high = "red", name = "Incident Count") +
  labs(title = "Heatmap of Race-on-Race Violence Incidents",
       x = "Perpetrator Race",
       y = "Victim Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x axis labels for better readability
```


## Figures 2-4

- **let's see if geography and victim are related**
- **For black perpetrators, it seems there's a heavier concentration in the south-east. Note: most unknown victim race crimes occur in the middle-south section**
- **For hispanic perpetrators, events occur in the mid-east. Uknown victim race crimes seem to be concentratedi n the north-east area.**
- **For white perpretrators, crime seems to be more spread out by victim race. It may be due to population density in specific areas.**
- **It appears that race and location have some correlation, let's test this later using Chi-Sq. independence** 


```{r viz2 cleanup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## zoom in on perp heatmap

black_perpetrator_data <- nypd_data %>%
  mutate(PERP_RACE = ifelse(PERP_RACE %in% c("WHITE HISPANIC", "BLACK HISPANIC"), "HISPANIC - MERGED", PERP_RACE),
         VIC_RACE = ifelse(VIC_RACE %in% c("WHITE HISPANIC", "BLACK HISPANIC"), "HISPANIC - MERGED", VIC_RACE)) %>%
  filter(PERP_RACE == "BLACK", !is.na(Longitude), !is.na(Latitude))  # Ensure location data is available

# Assuming that Latitude and Longitude need to be numeric
black_perpetrator_data$Latitude <- as.numeric(black_perpetrator_data$Latitude)
black_perpetrator_data$Longitude <- as.numeric(black_perpetrator_data$Longitude)

# Fetch a map of New York City
new_york_map <- get_map(location = 'New York City', zoom = 12)
```



```{r geomap, echo=FALSE, fig.cap="Heatmap: Geographic Distribution of Black perpetrated crimes vs. victim race"}
# Plotting the heatmap
ggmap(new_york_map) +
  geom_density2d(data = black_perpetrator_data, aes(x = Longitude, y = Latitude), size = 0.3) +
  stat_density2d(data = black_perpetrator_data, aes(x = Longitude, y = Latitude, fill = ..level.., alpha = ..level..),
                 geom = "polygon", color = NA) +
  scale_fill_gradient(low = "green", high = "red") +
  scale_alpha(range = c(0, 0.75), guide = FALSE) +
  facet_wrap(~VIC_RACE) +  # Create a separate map for each victim race
  labs(title = "Geographical Distribution of Black Perpetrator Incidents by Victim Race",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```



```{r viz2-cleanup-hispanic, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Consolidate Hispanic race categories and focus on Hispanic perpetrators
hispanic_perpetrator_data <- nypd_data %>%
  mutate(PERP_RACE = ifelse(PERP_RACE %in% c("WHITE HISPANIC", "BLACK HISPANIC"), "HISPANIC - MERGED", PERP_RACE),
         VIC_RACE = ifelse(VIC_RACE %in% c("WHITE HISPANIC", "BLACK HISPANIC"), "HISPANIC - MERGED", VIC_RACE)) %>%
  filter(PERP_RACE == "HISPANIC - MERGED", !is.na(Longitude), !is.na(Latitude))  # Ensure location data is available

# Ensure that Latitude and Longitude are numeric
hispanic_perpetrator_data$Latitude <- as.numeric(hispanic_perpetrator_data$Latitude)
hispanic_perpetrator_data$Longitude <- as.numeric(hispanic_perpetrator_data$Longitude)

# Fetch a map of New York City
new_york_map <- get_map(location = 'New York City', zoom = 12)



### Visualization: Heatmap for Geographic Distribution of Hispanic Perpetrated Crimes

```{r geomap-hispanic, echo=FALSE, fig.cap="Heatmap: Geographic Distribution of Hispanic Perpetrated Crimes vs. Victim Race"}
# Plotting the heatmap for Hispanic perpetrators
ggmap(new_york_map) +
  geom_density2d(data = hispanic_perpetrator_data, aes(x = Longitude, y = Latitude), size = 0.3) +
  stat_density2d(data = hispanic_perpetrator_data, aes(x = Longitude, y = Latitude, fill = ..level.., alpha = ..level..),
                 geom = "polygon", color = NA) +
  scale_fill_gradient(low = "green", high = "red") +
  scale_alpha(range = c(0, 0.75), guide = FALSE) +
  facet_wrap(~VIC_RACE) +  # Create a separate map for each victim race
  labs(title = "Geographical Distribution of Hispanic Perpetrator Incidents by Victim Race",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```




```{r viz2-cleanup-white, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Focus on White perpetrators
white_perpetrator_data <- nypd_data %>%
  filter(PERP_RACE == "WHITE", !is.na(Longitude), !is.na(Latitude))  # Ensure location data is available

# Ensure that Latitude and Longitude are numeric
white_perpetrator_data$Latitude <- as.numeric(white_perpetrator_data$Latitude)
white_perpetrator_data$Longitude <- as.numeric(white_perpetrator_data$Longitude)

# Fetch a map of New York City
new_york_map <- get_map(location = 'New York City', zoom = 12)
```

  

### Visualization: Heatmap for Geographic Distribution of White Perpetrated Crimes

```{r geomap-white, echo=FALSE, fig.cap="Heatmap: Geographic Distribution of White Perpetrated Crimes vs. Victim Race"}
# Plotting the heatmap for White perpetrators
ggmap(new_york_map) +
  geom_density2d(data = white_perpetrator_data, aes(x = Longitude, y = Latitude), size = 0.3) +
  stat_density2d(data = white_perpetrator_data, aes(x = Longitude, y = Latitude, fill = ..level.., alpha = ..level..),
                 geom = "polygon", color = NA) +
  scale_fill_gradient(low = "green", high = "red") +
  scale_alpha(range = c(0, 0.75), guide = FALSE) +
  facet_wrap(~VIC_RACE) +  # Create a separate map for each victim race
  labs(title = "Geographical Distribution of White Perpetrator Incidents by Victim Race",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```  
  
## Modeling -  chi-squared ind. test: race vs. location 

- **Given the heatmaps we've seen, let's try to understand if location has a high correlation with race of perpretrator**
- **Using chi-squared test for independence, we find a strong correlation in the existing data**
- **NOTE: data may be bias as we saw the unknowns are highly concentrated in one area**

```{r MODEL-CHISQ, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
nypd_data_boro <- nypd_data %>%
  filter(BORO != "UNKNOWN", !is.na(BORO), PERP_RACE != "UNKNOWN", !is.na(PERP_RACE))

contingency_table <- table(nypd_data_boro$BORO, nypd_data_boro$PERP_RACE)
chi_squared_result <- chisq.test(contingency_table)
print(chi_squared_result)

```

